---
title: "01_load"
author: Eric Torres, Lucía de Lamadrid, Konstantina Gkopi, Elena Iriondo and Jorge Santiago
format: html
editor: visual
---

### 1. Load packages

```{r}
#| message: false 
library(tidyverse) 
library(googledrive)
library(readr)
```

### 2. Download and read files

With the following code chunk, data is accessed programmatically from a link in Google Drive. The user has to register with their Google account in order to retrieve the data.

```{r}
# Definition of the folder in which we will store the datasets
raw_folder <- "../_raw"

# Check if the folder exists; create it if not
if (!dir.exists(raw_folder)) {
  dir.create(raw_folder)
  message("Created folder: ", raw_folder)
} else {
  message("Folder already exists: ", raw_folder)
}

# File ID on Google Drive
file_id <- "1CW63f5yjQsfovsbEJECvi1NlxkAIEer4"

# Definition of the path where it will be downloaded it
dest_path <- file.path(raw_folder, "MicrobiomeOTUtaxonomy.zip")

# Download the file into _raw
drive_download(as_id(file_id), path = dest_path, overwrite = TRUE)

# Unzip 
unzip(dest_path, exdir = raw_folder)
```

We import three datasets:

-   MicrobiomeWithMetadata.csv: it contains the relative abundance of the OTUs analysed per observation, as well as the conditions of the sample.

-   MicrobiomeMetadataDictionary.csv: it has the equivalence between the numerical values and the conditions.

-   MicrobiomeOTUtaxonomy.csv: it contains the taxonomic ranks of each analyses OTU.

```{r}
data_file_metadata <- "/data_zip/MicrobiomeWithMetadata.csv" 
data_file_dictionary <- "/data_zip/MicrobiomeMetadataDictionary.csv" 
data_file_otu <- "/data_zip/MicrobiomeOTUtaxonomy.csv"  

# Definition of the pathways of the datasets.
data_path_metadata = str_c(raw_folder, data_file_metadata) 
data_path_dictionary = str_c(raw_folder, data_file_dictionary) 
data_path_otu = str_c(raw_folder, data_file_otu) 
```

```{r}
#| message: false  
# Load data 
metadata_df <- read_csv(data_path_metadata) 
dictionary_df <- read_csv(data_path_dictionary)
otu_df <- read_csv(data_path_otu)  
```

In the `otu_df` there are two types of separators (`\t` followed by `,`) in the first column, so we will eliminate the `\t` to have a uniform data frame.

```{r}
otu_df <- otu_df |> 
  mutate(across(everything(), ~gsub("\t", "", .x)))
# change columns in all dataset taken from "https://stackoverflow.com/questions/43456687/using-dplyr-gsub-on-many-columns"
```

### 3. Create .tsv files

```{r}
data_folder = "../data"

if (!dir.exists(data_folder)) {
  dir.create(data_folder)
}

metadata_df |> 
  write_tsv(file = str_c(data_folder, "/01_data_metadata.tsv"))

dictionary_df |> 
  write_tsv(file = str_c(data_folder, "/01_data_dictionary.tsv"))

otu_df |> 
  write_tsv(file = str_c(data_folder, "/01_data_otu.tsv"))
```
