---
title: "Data augmentation"
author: Lucía de Lamadrid, Jorge Santiago Bajo, Eric Torres García
format:
  html:
    embed-resources: true
editor: visual
---

### 1. Load packages

```{r}
#| message: false 
#| echo: false
rm(list=ls())
library(tidyverse) 
library(readr)
```

### 2. Read data

```{r}
clean_df <- read.table('../data/02_metadata_long_filtered_label.tsv', header=TRUE, sep='\t')
head(clean_df)
```

In this section, we'll be working with the OTUs taxonomy file, so let's load it as well.

```{r}
otu_df_original <- read.table('../data/01_data_otu.tsv', header=TRUE, sep='\t')
head(otu_df_original)
```

The different columns in the OTU data are: Domain, Phylum, Class, Order, Family, Genus and Species. Kingdom is not represented in this case.

```{r}

```

If we do a deeper analysis of this dataset, we can actually check it is very ill-defined, if not directly erroneous. An example can be seen on OTU94, which presents several problems:

```{r}
otu_df |> 
  filter(OTU.ID == "OTU94")
```

The information from this OTU is the following:

-   Kingdom: Bacteria

-   Phyllum: Actinobacteria

-   Class: Actinobacteria

Here comes the first problems: they chose the same term for the phyllum and class! This is misleading and unnecessarily confusing. For disambiguate, the following synonyms could have been used instead: Actinomycetota for the phyllum and Actinomycetia for the class.

-   Order: Actinobacteridae

Actinobacteridae is actually a subclass, which is no longer in use (see <https://lpsn.dsmz.de/subclass/actinobacteridae>).

-   Family: Actinomycetales

Actinomycetales is actually an order, not a family.

-   Genus: Frankineae

Frankineae is an order, not a genus, and it is different from the previous order mentioned.

Additionally, there are two extra columns with no identification of taxonomy: Geodermatophilaceae and Blastococcus.

Such delirium of dataset is the result of complete lack of manual curation and care. Different actions need to be taken to make sense of the biological data.

```{r}
otu_df_modified <-  otu_df_original |> 
  select(OTU.ID, Phylum, Class) |> 
  mutate(Phylum = if_else(is.na(Phylum), "Unclassified", Phylum)) |> 
  mutate(Class = if_else(is.na(Class), "Unclassified", Class))

otu_df_modified
```

```{r}
clean_df <- clean_df |>  
  left_join(otu_df_modified, 
            join_by(OTU == OTU.ID)) |> 
  relocate(Phylum, Class, .after = OTU) 
```
