---
title: "04_describe"
author: Eric Torres, Lucia de Lamadrid, Konstantina Gkopi, Elena Iriondo and Jorge Santiago
format: html
editor: visual
---

## Data visualization and description

1.  Load packages

    ```{r}
    #| message: false 
    #| echo: false
    rm(list=ls())
    library(dplyr) 
    ```

2.  Read data

    ```{r}
    clean_df <- read.table('../data/03_metadata_long_filtered_taxonom.tsv', header=TRUE, sep='\t')
    head(clean_df)
    ```

    ```{r}
    raw_data <- read.table('../data/01_data_metadata.tsv', header=TRUE, sep='\t')
    head(raw_data)
    ```

### Abundance of OTUs in samples using phyla and class

```{r}
clean_df |> 
  filter(Diet %in% c("LFPP", "Western")) |> 
  group_by(Source, Diet) |> 
  mutate(norm_rel_abundance = rel_abundance / sum(rel_abundance)) |> # Normalize abundance within samples
  group_by(Source, Diet, Phylum) |> 
  summarise(mean_rel_abund = mean(norm_rel_abundance), .groups = "drop") |> # Calculate the mean relative abundances
  group_by(Source, Diet) |> 
  mutate(mean_rel_abund = mean_rel_abund / sum(mean_rel_abund)) |> # Normalize mean abundances so that the total for each one is 1
  ggplot(aes(x = Source,
             y = mean_rel_abund,
             fill = Phylum)) + 
  geom_bar(stat = "identity",
           position = "stack") +
  geom_hline(yintercept = 0) +
  facet_wrap(~ Diet) + 
  labs(
    title = "Normalized Relative Abundance of Phyla per source and diet type",
    x = "Source type",
    y = "Relative Abundance (Proportion)",
    fill = "Phyla"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", # change when data augmentation has been done
        legend.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

If we observe the above figure we can find some interesting insights. Firstly, we can find differences in phyla's relative abundance among the different parts of the mice's gut; but the differences are noteworthy between diet types, showing a lower abundance of *Proteobacteria* and higher *Verrucomicrobia* in Western diet compared to the LFPP diet.

```{r}
clean_df |> 
  filter(Diet %in% c("LFPP", "Western")) |> 
  group_by(Source, Diet) |> 
  mutate(norm_rel_abundance = rel_abundance / sum(rel_abundance)) |> # Normalize abundance within samples
  group_by(Source, Diet, Class) |> 
  summarise(mean_rel_abund = mean(norm_rel_abundance), .groups = "drop") |> # Calculate the mean relative abundances
  group_by(Source, Diet) |> 
  mutate(mean_rel_abund = mean_rel_abund / sum(mean_rel_abund)) |> # Normalize mean abundances so that the total for each one is 1
  ggplot(aes(x = Source,
             y = mean_rel_abund,
             fill = Class)) + 
  geom_bar(stat = "identity",
           position = "stack") +
  geom_hline(yintercept = 0) +
  facet_wrap(~ Diet) + 
  labs(
    title = "Normalized Relative Abundance of Class per source and diet type",
    x = "Source type",
    y = "Relative Abundance (Proportion)",
    fill = "Class"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", # change when data augmentation has been done
        legend.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Microbiota composition of mice with different diet and donor combination (relative abundance of phyla)

In order to compare the general composition of the microbiota in the mice with different donor and diet, we may represent the relative abundance of the most prevalent phyla.

For this purpose we will use the clean data.

```{r}
clean_df |> 
  filter(Diet %in% c("LFPP", "Western") & Donor %in% c("Fresh", "HMouseLFPP", "HMouseWestern") & Source != "Fresh" ) |> 
  group_by(Donor, Diet) |> 
  mutate(norm_rel_abundance = rel_abundance / sum(rel_abundance)) |> # Normalize abundance within samples
  group_by(Donor, Diet, Phylum) |> 
  summarise(mean_rel_abund = mean(norm_rel_abundance), .groups = "drop") |> # Calculate the mean relative abundances
  group_by(Donor, Diet) |> 
  mutate(mean_rel_abund = mean_rel_abund / sum(mean_rel_abund)) |> # Normalize mean abundances so that the total for each one is 1
  ggplot(aes(x = Donor,
             y = mean_rel_abund,
             fill = Phylum)) + 
  geom_bar(stat = "identity",
           position = "stack") +
  geom_hline(yintercept = 0) +
  facet_wrap(~ Diet) + 
  labs(
    title = "Normalized Relative Abundance of Phyla per donor and diet type",
    x = "Donor type",
    y = "Relative Abundance (Proportion)",
    fill = "Phylum"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", # change when data augmentation has been done
        legend.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Biodiversity of humanised mice microbiota with different diet

We can use diversity indices in order to quantify biodiversity. As we know the relative abundance of each OTU, we can use the Shannon diversity index. Based on Claude Shannon's formula for entropy, this popular ecological metric takes into account the number of species living in a habitat (richness) and their relative abundance (evenness).

The formula we will use is:

$$
H' = -\sum_{i=1}^R p_i \ln p_i \\
p_i \text{ is the relative abundance of OTU}_i\\ R \text{ is the total number of OTUs}
$$

We will compare this index for first generation humanised mice following each diet. For this purpose, we will use the original raw data, in which all OTU counts (including the less abundant ones) are included.

We first store the diet column in a dataframe.

```{r}
diet_data <- raw_data |> 
  filter(Donor == 3 & Source != 4 & Diet %in% c(0, 1)) |> #we select 1st generation humanised mice (donor is fresh), get rid of microbiota samples that were not taken at the time of sacrifice (source is not fecal), and we select the LFPP (0) and Western (1) diets
  select(Diet) |> 
  mutate(Diet = case_when(Diet == 0 ~ "LFPP",
                          Diet == 1 ~ "Western"))
```

We calculate the Shannon index for each row and save it in a new dataframe.

```{r}
shannon <- raw_data |> 
  filter(Donor == 3 & Source != 4 & Diet %in% c(0, 1)) |> #same filters as before
  select(starts_with("OTU")) |> #we only need the OTU columns to calculate the index 
  mutate(across(everything(), ~ (-1)*.x * log(.x))) |> #we transform each OTU relative abundance column (p_i) into p_i*log(p_i) 
  rowwise() |> #we apply the following operations row by row
  mutate(
    sum_all_columns = sum(across(everything()), na.rm = TRUE), #we add the products to calculate the shannon_index
    .keep = "unused") |> #we only keep the new shannon index column
  ungroup()
shannon
```

We combine both dataframes and calculate the Shannon index and its standard deviation for each diet.

```{r}
combined_df <- bind_cols(diet_data, shannon)  #combination of both dataframes
final_shannon <- combined_df |> #generate a final data frame with the shannon index and sd for each diet
  group_by(Diet) |>
  summarise(
    shannon_index= mean(sum_all_columns),
    standard_dev = sd(sum_all_columns)
  )
final_shannon
```

Finally, we plot the index for each of the studied diets.

```{r}
# Plot with error bars
ggplot(final_shannon, aes(x = Diet, y = shannon_index)) +
  geom_col(aes(fill=Diet)) +
  geom_errorbar(aes(ymin = shannon_index - standard_dev, ymax = shannon_index + standard_dev), width = 0.2) +  # Error bars
  labs(
    title = "Biodiversity in microbiota",
    x = "Diet",
    y = "Shannon diversity index"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Interestingly, the Western diet leads to a more biodiverse microbiota than the "healthier" LF-PP diet in the humanised mice.
