---
title: "06_analysis_2"
author: Konstantina Gkopi
format: html
editor: visual
---

### PCA Analysis on Phylum-Level Aggregated Microbiome Data

In this analysis, we perform PCA on microbiome data aggregated at the phylum level. We aim to identify key axes of variation and visualize sample clustering to uncover patterns in microbial community composition.

#### 1. **Load libraries**

```{r}
library(tidyverse)
library(broom)
library(ggrepel)
library(cowplot)
```

#### 2. Load data

For this analysis we will use the "03_metadata_long_filtered_taxonom.tsv" data frame, that contains metadata, OTU counts and taxonomic information for each sample.

```{r}
# load data
clean_df <- read_tsv("../data/03_metadata_long_filtered_taxonom.tsv")
head(clean_df)
```

#### 3. Prepare data for PCA

To perform PCA, we aggregate the microbiome data by summing relative abundances within each phylum. This reduces dimensionality while retaining relevant biological information. We then reshape the data into a wide format suitable for PCA analysis.

```{r}
# Aggregate relative abundances by phylum
aggregated_data <- clean_df |> 
  group_by(SampleID, Donor, Phylum) |> 
  summarize(rel_abundance = sum(rel_abundance), .groups = "drop")

# Pivot wider to prepare for PCA (SampleID by Phylum)
aggregated_wide <- aggregated_data |> 
  pivot_wider(names_from = Phylum, values_from = rel_abundance, values_fill = 0)

# Check of the aggregated data
head(aggregated_wide)
```

#### 4. Performing PCA

Using the aggregated data, we scale the phylum-level abundances and apply PCA. This identifies the key axes of variation in the dataset, summarizing complex microbial community structures into principal components.

```{r}
# Scale the aggregated data and perform PCA
otu_data_pca <- aggregated_wide |> 
  select(-SampleID, -Donor) |> 
  scale() |> 
  prcomp()
```

#### 5. Explaining Variance

To understand how much of the dataset’s variability is captured by each principal component, we analyze the explained variance.

```{r}
pca_variance <- otu_data_pca |>
  tidy(matrix = "eigenvalues")

pca_variance
```

```{r}
# Plot variance explained by PCs
ggplot(pca_variance, aes(x = PC, y = percent)) +
  geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7, width = 0.8) +
  geom_line(aes(y = cumulative), color = "red", size = 1.2, group = 1) +
  geom_point(aes(y = cumulative), color = "red", size = 2) +
  labs(
    title = "Variance Explained by Top Principal Components",
    x = "Principal Component",
    y = "Percentage of Variance Explained"
  ) +
  scale_x_continuous(breaks = 1:7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank()
  )
```

The PCA variance table shows that PC1 and PC2 together explain 44.07% of the total variance, with PC1 accounting for 26.23% and PC2 for 17.84%. Including PC3 increases the cumulative explained variance to 59.61%, capturing most of the dataset’s variability. Variance contributions drop significantly for subsequent components, making PC1 and PC2 the most relevant for analysis. This highlights the effectiveness of PCA in summarizing key microbial variation driven by donor origin, diet, and phylum-level differences.

#### 6. **Visualizing PC1 vs PC2**

We plot the first two principal components (PC1 and PC2) to visualize the differences between donors.

```{r}
# Visualize PC1 VS PC2
otu_data_pca |> 
  augment(aggregated_wide) |> 
  ggplot(aes(.fittedPC1, .fittedPC2, fill = Donor)) +
  geom_point(size = 1.5, shape = 21) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "PCA: PC1 vs PC2 by Donor Group",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank()
  )
```

```{r}
# Visualize PC1 VS PC2
otu_data_pca |> 
  augment(aggregated_wide) |> 
  ggplot(aes(.fittedPC1, .fittedPC2, fill = Donor)) +
  geom_point(size = 1.5, shape = 21) +
  facet_wrap(~ Donor) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "PCA: PC1 vs PC2 by Donor Group",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank()
  )
```

#### 7. **Understanding contributions to PC1 and PC2**

The contributions of individual phyla to the first two principal components are analyzed to identify key taxa driving variation in the data. This provides biological insights into which phyla influence microbiome differences.

```{r}
# Define arrow style for plotting
arrow_style <- arrow(
  angle = 25, ends = "first", type = "closed", unit(0.2, "cm")
)
rotation_matrix <- otu_data_pca |> tidy(matrix = "rotation")

rotation_pivot <- rotation_matrix |>
  filter(PC %in% c(1, 2)) |>
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value")

# Print to confirm the structure
print(rotation_pivot)

ggplot(rotation_pivot, aes(x = PC1, y = PC2, color = column)) +
  geom_segment(aes(xend = 0, yend = 0), arrow = arrow_style, size = 0.8) +
  scale_color_brewer(palette = "Dark2") +
  geom_text_repel(aes(label = column), size = 3, hjust = 1) +
  labs(
    title = "Contributions of Phyla to PC1 and PC2",
    x = "PC1 Contributions",
    y = "PC2 Contributions",
    color = "Phylum"
  ) +
  coord_fixed() +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "none",
    panel.grid.major = element_blank()
  )
```

The biplot illustrates the contributions of different phyla to the first two principal components (PC1 and PC2). **Bacteroidetes** and **Firmicutes** contribute strongly to PC1 in opposite directions, highlighting their contrasting roles in shaping microbiota variability. **Verrucomicrobia** and **Unclassified** taxa influence PC2 significantly, reflecting unique microbial variability in specific groups. Minor contributions from **Proteobacteria**, **Actinobacteria**, and **TM7** suggest their subtler impact on overall microbiota structure. This plot emphasizes that a few dominant phyla drive the main axes of variation in microbial communities.

### Analysis of Microbiome Clusters by Donor Groups Using Hierarchical Clustering

We decided to conduct hierarchical clustering because PCA, while effective at reducing dimensionality and visualizing overall variability, was not sufficient to capture distinct groupings and the relationships between donor origins and microbiota compositions. PCA provides a global overview of variation but does not explicitly group samples into discrete clusters.

Hierarchical clustering, on the other hand, allowed us to group microbiota profiles into well-defined clusters based on similarity, making it easier to interpret patterns within and across donor groups. By combining this with statistical tests (e.g., chi-squared test), we could confirm the significance of the associations between donor origins and microbiota clusters, providing a more detailed understanding of the data.

#### 1. Data Preparation

For this analysis we will use the 02_metadata_wide_filtered_label.tsv" data frame, that contains metadata and OTU counts for each sample.

```{r}
# Read the metadata in wide format
filtered_metadata_wider <- read_tsv("../data/02_metadata_wide_filtered_label.tsv")

# Select OTU columns
otu_data <- filtered_metadata_wider |>
  select(starts_with("OTU"))

# Check the dimensions of the OTU data
dim(otu_data)

```

#### **Standardization**

To ensure clustering algorithms work effectively, we standardize the OTU data. This ensures all features are on the same scale.

```{r}
# Scale the OTU data
otu_data_scaled <- otu_data |> 
  scale()

# Convert scaled matrix back to tibble for tidyverse compatibility
otu_data_scaled <- as_tibble(otu_data_scaled)

```

We add the relevant metadata, in this case "Donor", as we will use it later in our analysis.

```{r}
# Add relevant metadata (e.g., Donor)
otu_data_with_metadata <- otu_data_scaled |>
  mutate(Donor = filtered_metadata_wider |> pull(Donor))
```

#### 2. Hierarchical Clustering

We compute a **distance matrix** using the Euclidean distance to measure similarity between microbiome samples. The clustering is performed using Ward’s method, which minimizes variance within clusters.

```{r}
# Compute Euclidean distance matrix
dist_matrix <- otu_data_scaled |>
  dist()
# Perform hierarchical clustering
hclust_result <- hclust(dist_matrix, method = "ward.D2")
```

#### **Cluster Assignment**

We cut the dendrogram into three clusters and assign cluster labels to the metadata. This step groups samples based on the similarity of their OTU profiles.

```{r}
# Cut dendrogram into 3 clusters
cluster_labels <- cutree(hclust_result, k = 3) |>
  as_tibble() |>
  rename(Cluster = value)

# Attach cluster labels to metadata
clustered_metadata <- filtered_metadata_wider |> 
  mutate(Cluster = cluster_labels |> pull(Cluster))

```

#### 3. **Donor Distribution Across Clusters**

To understand the relationship between clusters and donor groups, we summarize the distribution of donors within each cluster. This helps identify patterns of donor contributions.

```{r}
# Summarize cluster-donor relationships
cluster_metadata_summary <- clustered_metadata |>
  group_by(Cluster, Donor) |>
  summarize(Count = n(), .groups = "drop") |>
  group_by(Cluster) |>
  mutate(Proportion = Count / sum(Count))

# View the summary
cluster_metadata_summary
```

#### **Visualization**

We use a bar plot to visualize the proportion of donor groups within each cluster. This reveals how donors contribute to the composition of each cluster.

```{r}
# Bar plot showing donor proportions within each cluster
ggplot(cluster_metadata_summary, aes(x = Cluster, y = Proportion, fill = Donor)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Proportion of Donor Groups Within Clusters",
    x = "Cluster",
    y = "Proportion",
    fill = "Donor"
  ) +
  theme_minimal()

```

#### 4. **Statistical Association Between Donor and Cluster**

We simplify the donor groups by grouping less frequent donors under "Other" to streamline the analysis. A contingency table is created to summarize the relationship between donor groups and clusters.

```{r}
# Simplify donor groups and create a contingency table
clustered_metadata <- clustered_metadata |>
  mutate(Donor = case_when(
    Donor %in% c("Human", "HMouseWestern", "CONVD" ) ~ "Other",
    TRUE ~ Donor
  ))
donor_cluster_table <- clustered_metadata |>
  count(Cluster, Donor) |> 
  pivot_wider(names_from = Donor, values_from = n, values_fill = 0) |> 
  column_to_rownames("Cluster") |> 
  as.matrix()
```

#### **Chi-Squared Test**

A chi-squared test is performed on the contingency table to determine whether donor distributions across clusters are statistically significant.

```{r}
# Perform chi-squared test
chi2_result <- chisq.test(donor_cluster_table)
chi2_result
```

The hierarchical clustering results reveal distinct donor distributions. **Cluster 1** is dominated by **HMouseLFPP** (55.5%) with notable contributions from **Frozen** (17.8%) and **Fresh** (18.7%), reflecting plant-rich diets and preserved samples. **Cluster 2** includes mostly **Fresh** (55.1%) and **HMouseLFPP** (26.5%), indicating a mix of human-derived and dietary influences. **Cluster 3** is almost entirely **CONVR** (95%), representing natural microbiota from control mice. The chi-squared test (p \< 2.2e-16) confirms significant associations between donor origins and clusters, highlighting the influence of donors on microbiota composition.
